Precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-14 manuscript.tex
Update on Overleaf.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONTEXT
Neural activity is directly influenced by our immediate environment and by internal states and is structured in order to generate motor actions. The efficiency of these actions is key for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity, as exemplified by the recordings observed in neurobiological experiments. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation \add[Antoine]{of cognitive processes} within the neural activity. Nevertheless, we will use this terminology in all generality to denote the existence of a structure in neural activity. In this respect, it is \change[Antoine]{known}{reasonable to declare} that neural activity may be related to specific measurable variables\change[Antoine]{, and since Galvani's}{. Since Galvani's} experiments, we know that an electrical activity applied on muscular nerves can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A central and well-studied way of communication between neurons is specific electrochemical events called action potentials, or spikes, which were first discovered \change[Antoine]{by Lord Adrian}{at the beginning of the $XX^{th}$ century}~\citep{adrian_impulses_1926}.
In the scope of this article, we focus mainly on these spikes, which are brief (about one millisecond) impulses that propagate along the axons of neurons. First, these have the particular property of being essentially binary in their amplitude, i.e. they are prototypical, 
``\href{https://en.wikipedia.org/wiki/All-or-none_law}{all-or-none}''. \change[Antoine]{Typically, these are received by a neuron from other afferent neurons on the arborized ``input'' dendrite, integrated in its soma, and possibly further propagated along its ``output'' axon to reach efferent neurons.}{Typically, a postsynaptic neuron receives incoming spikes from other afferent neurons on the arborized ``input'' dendrite. The integration of these spikes by the dendritic tree and the soma of the postsynaptic neuron results in the modification of its membrane potential that possibly leads to the emission of an action potential along its ``output'' axon to reach efferent neurons.}
The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. \change[Antoine]{Indded, neurophysiologists typically}{Most neurophysiologists} use the temporal evolution of the firing rate (for instance, as computed as the average occurrence of spikes in small temporal windows of about $100~\ms$) in order to characterize the dynamical activity of neurons. This may be extended by computing different statistics on each neuron's sequence of spikes~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.

@@

In comparison to a classical \change[Antoine]{analogical}{analog} vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of representations \change[Antoine]{(which rely on the firing rate)}{that rely on the firing rate} by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004,dilorenzo_spike_2013}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %

@@

Although surprising, this speed of the visual cortex in primates is compatible with the latencies that are recorded at the neuro-physiological level. The rapid propagation of the visual information in the thalamus, then in the primary visual cortex takes about $45~\ms$ in the macaque~\citep{schmolesky_signal_1998} and about $60~\ms$ in humans~\citep{vanni_coinciding_2001}. This functioning of visual processing as a forward pass is most prominent in fast processing (see Figure~\ref{fig:thorpe}), and can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}. An important consequence of this speed of processing of vision is that it implies that processing is carried out using only very few spikes per layer. As a comparison, the latencies in macaque monkeys are about: Retina, $20-40~\ms$; V1, $40-60~\ms$; IT, $80-100~\ms$; MC, $140-190~\ms$; to finger muscles, $180-260~\ms$. Indeed, we may consider that a behavioral response occurring in only $200~\ms$ would involve about ten processing stages along the ``forward'' pathways of the visual system. Such models were indeed efficiently modeled in feed-forward models trained with back-propagation~\citep{serre_feedforward_2007,jeremie_ultrafast_2022}. At the same time, it has been demonstrated that one spike requires a significant amount of time (about $10~\ms$) to be conducted from one layer to the next~\citep{nowak_timing_1997,thorpe_seeking_2001}.  This figure is inspired by similar schematics done for monkeys in~\citep{thorpe_seeking_2001}. As a consequence, these results suggest that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How precise spike timing may encode vectors of real values}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% reliability of spike timings
Sensory data is most often represented by continuous values, such as the energy produced by a flow of photons that hits the different photoreceptors of the retina. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate of cortical cells in awake monkeys is highly irregular~\citep{softky_highly_1993}, which makes it at first sight inconsistent with the temporal integration of firing rate. Remarkably, it has been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible: while the first spike is aligned to stimulus' time, the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{bryant_spike_1976,mainen_reliability_1995} (see Figure~\ref{fig:mainen}). However, if that same neuron is now driven by a \emph{frozen} noise, that is, a highly dynamic signal which is repeatedly replayed from trial to trial, then the output spikes are highly reproducible (for a review, see~\citep{ermentrout_reliability_2008}). This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

% retina
How may information about the sensory world be encoded in the spiking pattern? At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.
%----------------------------%
% C_MainenSejnowski1995_Perrinet.ipynb
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{\textbf{Reproducibility of the spiking response of a neuron. }The timing of the spikes produced following the repetition of a step stimulus is less reproducible than that to a noisy stimulus. The stimulus current value over time for a step stimulus (top left) and for a noisy one (top right). Trial repetitions of a Leaky Integrate-and-Fire neuron stimulated by the stimulus on the upper row (middle row). Membrane potential is represented by dark blue color when light yellow colors when depolarized) and quantified by . The average firing rate across trials (lower row). While this seems paradoxical at first sight, it highlights the consequence of using the same \emph{frozen} noise at each repetition and highlights the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this~\href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995_Perrinet.ipynb}{notebook} for a replication of the results from~\citep{mainen_reliability_1995}  using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

% %----------------------------%
% \begin{figure}
% \centering
% \note{redo a figure similar to the thorpe code}%\includegraphics{images/roc.jpg}
% \caption{Latency coding. An input analog profile is encoded in latencies: the higher the contrast, the shorter the latency. In this example, one generates at most one spike per neuron.
% }\label{fig:roc}
% \end{figure}
% %----------------------------%
In fact, similar results have been demonstrated through neurophysiological recordings in the primary visual cortex and show that different levels of visual activity will induce different levels of neuronal discharge latency in the primary visual area~\citep{celebrini_dynamics_1993}. First-spike latency codes are a feasible mechanism for information transfer, even when biologically plausible estimates of stimulus onset are considered, for instance for sound localization~\citep{chase_first-spike_2007}.  Note also that timing is not entirely sensory or internal, but can be used as a general neural coding principle. In~\citep{safaie_turning_2020} for instance, they found that ``timing accuracy was improved when the environment afforded cues that rats can incorporate into motor routines. Timing, at least in animals, may thus be fundamentally embodied and situated.''  Many models have used these properties in temporal coding to build fast image categorization networks~\citep{gautrais_rate_1998,delorme_spikenet_1999, perrinet_coding_2004}. These models take the form of artificial spiking neural networks (SNNs) and have been able to demonstrate their practical applications for image categorization~\citep{delorme_ultra-rapid_2000}. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron by using the rank of neurons' activation~\citep{delorme_spikenet_1999,bonilla_analyzing_2022}. Another class of artificial SNNs uses precise spike timing as a metric in order to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been extended in novel gradient-based methods. The subsequent surrogate gradient method is now widely used in methods that attempt to transfer performance from analog (CNNs) to spike-based (SNNs) architectures~\citep{zenke_remarkable_2021}. This type of modeling often uses the classical task of categorizing images developed in deep learning, while adapting it to the specificity of the event-based representation~\citep{goltz_fast_2021}. For instance,~\citep{kheradpisheh_stdp-based_2018} implements a STDP-based spiking deep convolutional neural networks for object recognition or~\citep{tavanaei_representation_2018} develops a form of spike-based, competitive learning applied for unsupervised learning.

However, the performance of SNNs is still lagging compared to that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open. \change[Antoine]{Achieving to efficiently implement this type of modeling is extremely important with respect to the development of a new generation of event-based cameras}{Improvements in this new generation of Artificial Neural networks (ANN) would bring major advances in terms of efficient computations in machine learning. They would benefit in particular to a new generation of cameras called Silicon Retinas}~\citep{gallego_event-based_2022} (see Section~\ref{sec:neuromorphic}). %These represent visual information as events, similarly to the biologically-inspired representation described above and in contrast with a basic frame-based representation~\citep{lichtsteiner_128x128_2008} (for a review, see~\citep{rasetto_challenges_2022}, this will be developed in Section~\ref{sec:neuromorphic}).% This work has been extended to include unsupervised learning capabilities, and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{grimaldi_homeostatic_2021,grimaldi_robust_2022}. 
To alleviate this gap in computational neuroscience, we will further review the role of  precise spike timing in neural assemblies from a neuroscientific perspective in the next section.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Role of precise spike timing in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One first hypothesis: synchronous firing in cell assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that neural assemblies could form so-called ``synfire chains,'' that is, showing the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. More broadly, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.

>>>>

\begin{document}
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
\subsection{Is there a neural code?}
% CONTEXT
Neural activity is directly influenced by our immediate environment and by internal states and is structured in order to generate motor actions. The efficiency of these actions is key for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity, as exemplified by the recordings observed in neurobiological experiments. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation \add[Antoine]{of cognitive processes} within the neural activity. Nevertheless, we will use this terminology in all generality to denote the existence of a structure in neural activity. In this respect, it is \change[Antoine]{known}{reasonable to declare} that neural activity may be related to specific measurable variables\change[Antoine]{, and since Galvani's}{. Since Galvani's} experiments, we know that an electrical activity applied on muscular nerves can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A central and well-studied way of communication between neurons is specific electrochemical events called action potentials, or spikes, which were first discovered \change[Antoine]{by Lord Adrian}{at the beginning of the $XX^{th}$ century}~\citep{adrian_impulses_1926}. In this study, the frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb.
In the scope of this article, we focus mainly on these spikes\change[Antoine]{which are brief (about one millisecond) impulses that propagate along the axons of neurons. First, these have the particular property of being essentially binary in their amplitude, i.e. they are prototypical, ``all-or-none''}{in vertebrate systems. They can be described as brief (about one millisecond) and prototypical, i.e. ``all-or-none'', impulses that propagate along the axons of neurons.}
 \change[Antoine]{Typically, action potentials are received by a neuron from other afferent neurons on the arborized ``input'' dendrite, integrated in its soma, and possibly further propagated along its ``output'' axon to reach efferent neurons.}{Typically, a postsynaptic neuron receives incoming spikes from other afferent neurons on the arborized ``input'' dendrite. The integration of these spikes by the dendritic tree and the soma of the postsynaptic neuron results in the modification of its membrane potential that possibly leads to the emission of an action potential along its ``output'' axon to reach efferent neurons.} \add[Antoine]{Except in the retina where neurons communicate with graded potentials}~\citep{goudas_graded_1960}\add[Antoine]{, it is assumed that spike trains are the main component of the neural code.}
\remove[Antoine]{The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb(Adrian citation), and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code.} \change[Antoine]{Indded, neurophysiologists typically}{Until recently, most neurophysiologists} use the temporal evolution of the firing rate (for instance, as computed as the average occurrence of spikes in small temporal windows of about $100~\ms$) in order to characterize the dynamical activity of neurons. This may be extended by computing different statistics on each neuron's sequence of spikes~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.

@@

In comparison to a classical \change[Antoine]{analogical}{analog} vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of representations \change[Antoine]{(which rely on the firing rate)}{of the rate coding hypothesis that relies only on the firing rate} by considering \change[Antoine]{a representation based on repetitions of spiking motifs at precise times of occurrence.}{representations based on the precise timing of single spikes.} \remove[Antoine]{This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence.} Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004,dilorenzo_spike_2013}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons. \add[Antoine]{This review } A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %

@@

Although surprising, this speed of the visual cortex in primates is compatible with the latencies that are recorded at the neuro-physiological level. The rapid propagation of the visual information in the thalamus, then in the primary visual cortex takes about $45~\ms$ in the macaque~\citep{schmolesky_signal_1998} and about $60~\ms$ in humans~\citep{vanni_coinciding_2001}. This functioning of visual processing as a forward pass is most prominent in fast processing (see Figure~\ref{fig:thorpe}), and can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}. An important consequence of this speed of processing of vision is that it implies that processing is carried out using only very few spikes per layer. As a comparison, the latencies in macaque monkeys are about: Retina, $20-40~\ms$; V1, $40-60~\ms$; IT, $80-100~\ms$; MC, $140-190~\ms$; to finger muscles, $180-260~\ms$. Indeed, we may consider that a behavioral response occurring in only $200~\ms$ would involve about ten processing stages along the ``forward'' pathways of the visual system. Such processes were indeed efficiently reproduced in feed-forward models trained with back-propagation~\citep{serre_feedforward_2007,jeremie_ultrafast_2022}. At the same time, it has been demonstrated that one spike requires a significant amount of time (about $10~\ms$) to be conducted from one layer to the next~\citep{nowak_timing_1997,thorpe_seeking_2001}.  This figure is inspired by similar schematics done for monkeys in~\citep{thorpe_seeking_2001}. As a consequence, these results suggest that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How precise spike timing may encode vectors of real values}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% reliability of spike timings
Sensory data is most often represented by continuous values, such as the energy produced by a flow of photons that hits the different photoreceptors of the retina. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate of cortical cells in awake monkeys is highly irregular~\citep{softky_highly_1993}, which makes it at first sight inconsistent with the temporal integration of firing rate. Remarkably, it has been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible: while the first spike is aligned to stimulus' time, the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{bryant_spike_1976,mainen_reliability_1995} (see Figure~\ref{fig:mainen}). However, if that same neuron is now driven by a \emph{frozen} noise, that is, a highly dynamic signal which is repeatedly replayed from trial to trial, then the output spikes are highly reproducible (for a review, see~\citep{ermentrout_reliability_2008}). This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

% retina
\remove[Antoine]{How may information about the sensory world be encoded in the spiking pattern?} At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.
%----------------------------%
% C_MainenSejnowski1995_Perrinet.ipynb
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{\textbf{Reproducibility of the spiking response of a neuron. }The timing of the spikes produced following the repetition of a step stimulus is less reproducible than that to a noisy stimulus. The stimulus current value over time for a step stimulus (top left) and for a noisy one (top right). Trial repetitions of a Leaky Integrate-and-Fire neuron stimulated by the stimulus on the upper row (middle row). Membrane potential is represented by dark blue color when light yellow colors when depolarized) and quantified by . The average firing rate across trials (lower row). While this seems paradoxical at first sight, it highlights the consequence of using the same \emph{frozen} noise at each repetition and highlights the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this~\href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995_Perrinet.ipynb}{notebook} for a replication of the results from~\citep{mainen_reliability_1995}  using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

% %----------------------------%
% \begin{figure}
% \centering
% \note{redo a figure similar to the thorpe code}%\includegraphics{images/roc.jpg}
% \caption{Latency coding. An input analog profile is encoded in latencies: the higher the contrast, the shorter the latency. In this example, one generates at most one spike per neuron.
% }\label{fig:roc}
% \end{figure}
% %----------------------------%
In fact, similar results have been demonstrated through neurophysiological recordings in the primary visual cortex and show that different levels of visual activity will induce different levels of neuronal discharge latency in the primary visual area~\citep{celebrini_dynamics_1993}. First-spike latency codes are a feasible mechanism for information transfer, even when biologically plausible estimates of stimulus onset are considered, for instance for sound localization~\citep{chase_first-spike_2007}.  Note also that timing is not entirely sensory or internal, but can be used as a general neural coding principle. In~\citep{safaie_turning_2020} for instance, they found that ``timing accuracy was improved when the environment afforded cues that rats can incorporate into motor routines. Timing, at least in animals, may thus be fundamentally embodied and situated.''  Many models have used these properties in temporal coding to build fast image categorization networks~\citep{gautrais_rate_1998,delorme_spikenet_1999, perrinet_coding_2004}. These models take the form of artificial spiking neural networks (SNNs) and have been able to demonstrate their practical applications for image categorization~\citep{delorme_ultra-rapid_2000}. One of these is the SpikeNet algorithm, which uses a purely temporal approach by encoding information using one spike per neuron by using the rank of neurons' activation~\citep{delorme_spikenet_1999,bonilla_analyzing_2022}. Another class of artificial SNNs uses precise spike timing as a metric in order to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been extended in novel gradient-based methods. The subsequent surrogate gradient method is now widely used in methods that attempt to transfer performance from analog (CNNs) to spike-based (SNNs) architectures~\citep{zenke_remarkable_2021}. This type of modeling often uses the classical task of categorizing images developed in deep learning, while adapting it to the specificity of the event-based representation~\citep{goltz_fast_2021}. For instance,~\citep{kheradpisheh_stdp-based_2018} implements a STDP-based spiking deep convolutional neural networks for object recognition or~\citep{tavanaei_representation_2018} develops a form of spike-based, competitive learning applied for unsupervised learning. 
However, the performance of SNNs is still lagging compared to that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open. \change[Antoine]{Achieving to efficiently implement this type of modeling is extremely important with respect to the development of a new generation of event-based cameras}{Improvements in this new generation of Artificial Neural networks (ANN) would bring major advances in terms of efficient computations in machine learning. They would benefit in particular to a new generation of cameras called Silicon Retinas}~\citep{gallego_event-based_2022} (see Section~\ref{sec:neuromorphic}). %These represent visual information as events, similarly to the biologically-inspired representation described above and in contrast with a basic frame-based representation~\citep{lichtsteiner_128x128_2008} (for a review, see~\citep{rasetto_challenges_2022}, this will be developed in Section~\ref{sec:neuromorphic}).% This work has been extended to include unsupervised learning capabilities, and we have recently developed a SNN architecture that allows to categorize images of different classes in only a few spikes~\citep{grimaldi_homeostatic_2021,grimaldi_robust_2022}.

\change[Antoine]{To alleviate this gap in computational neuroscience, we will further review the role of  precise spike timing in neural assemblies from a neuroscientific perspective in the next section.}{Even if technology lags far behind biology, this introduction demonstrates the importance of timing in neural processes and we will further review the role of precise spike timing in neural assemblies. We start by reviewing the different hypotheses that aim at deciphering the neural code with spatio-temporal spiking motifs. After listing some biological evidence for the use of precise spike timing we review some computational models and neuromorphic technics that add this temporal dimensions to their computations.}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Role of precise spike timing in neural assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{One first hypothesis: synchronous firing in cell assemblies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that neural assemblies could form so-called ``synfire chains,'' that is, showing the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. More broadly, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.
